Parametricity Consequences of GADTs Representations

### 2021-02-23 20-gadts/GADT-story.tex
work on gadt-story

GADTs are traditionally represented via Church encodings. In this
view, GADTs are completely determined by their syntax; for example,
the GADT $\mathsf{G}$ defined in Haskell by
\begin{equation}\label{eq:seq}
\begin{array}{l}
\mathsf{data\, Seq\,a\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Seq \,a \to Seq\,b \to
  Seq\,(a \times b)}\\ 
\end{array}
\end{equation}
\noindent
comprises sequences of any type $\mathsf{a}$ and sequences obtained by
pairing the data in two already existing sequences. By contrast, the
primitive representation sees a GADT as the {\em functorial
  completion}~\cite{jp19} of its syntax. Completing a GADT's syntax to
a {\em functor} is necessary in order to be able to take the least
fixed point~\cite{tfca}, so it is inherent in the view of data types
as their primitive representations. Viewed as its primitive
representation, $\mathsf{Seq}$ comprises not just those data elements
representable in syntax, but also all data elements of the form
$\mathsf{map\,f\,s}$ for all types $\mathsf{t_1}$ and $\mathsf{t_2}$,
all functions $\mathsf{f : t_1 \to t_2}$ definable in the language,
and all $\mathsf{s :: Seq\,t_1}$, as well as all data elements of the
form $\mathsf{map\,g\,s'}$ for each appropriately typed function
$\mathsf{g}$ and each element $\mathsf{s'}$ already added to the data
type, and so on. Functorial completion for $\mathsf{Seq}$ adds, in
particular, data elements of the form
$\mathsf{map\,g\,(Pair\,u_1\,u_2)}$ even though these may not
themselves be of the form $\mathsf{Pair\,v_1\,v_2}$ for any terms
$\mathsf{v_1}$ and $\mathsf{v_2}$. By contrast, functorial completion
adds no new data elements to the syntax of a GADT that is actually a
(polynomial) ADT or a (hereditarily polynomial) nested type, so its
Church encoding and primitive representation coincide. However, these
representations are rarely the same for GADTs.

@@

Requiring $F$ to be a functor will ensure that the interpretation $\mu
F$ of the ADT $\mathsf{T\,a} = \mathsf{\mu F}$ exists. But to ensure
that $\mu F$ is itself a functor, so that the type constructor
$\mathsf{T}$ associated with the ADT $\mathsf{T\,a}$ supports a
$\mathsf{map}$ function as well, we can require that $F$ also be the
fixpoint $\mu H$ of a functor $H$. In this case, however, $H$ must be
a functor on the category $\mathit{Set}^\mathit{Set}$ of functors and
natural transformations on $\mathit{Set}$. That is, $H$ must be a {\em
  higher-order} functor on $\mathit{Set}$. Reflecting this requirement
back into syntax gives that $\mathsf{T\,a} = \mathsf{(\mu F)\,a} =
\mathsf{(\mu\,(\mu H))\,a}$ for the ``type constructor constructor''
$\mathsf{H}$ that supports suitable\footnote{The $\mathsf{map}$
  function for $\mathsf{H}$ should satisfy syntactic reflections of
  the functoriality properties in $\mathit{Set}^\mathit{Set}$ ---
  i.e., preservation of identity natural transformations and
  composition of natural transformations --- and the $\mathsf{map}$
  function for $\mathsf{H\,F}$ should satisfy syntactic reflections of
  the functoriality properties in $\mathit{Set}$ as in the previous
  footnote. Again, there is no mechanism in Haskell for enforcing
  this.} $\mathsf{map}$ functions.

@@

A concrete example is given by the ADT $\mathsf{List\,a}$. This type
is modelled as the fixpoint $\mu F_{\mathit{List\,a}}$ of the
first-order functor whose action on sets is given by
$F_{\mathit{List}\,a}\,X = 1 + a \times X$ and whose action on
functions is given by $F_{\mathit{List}\,a}\,f = 1 + \id_a \times
f$. The type constructor $\mathsf{List}$ is modelled by the functor
that is the fixpoint $\mu H$ of the higher-order functor $H$ whose
action on a functor $F$ is given by the functor $H\,F$ whose actions
on sets and functions between them are given by $H\,F\,X = 1 + X
\times F\,X$ and $H\,F\,f = \id_1 + f \times F\,f$, respectively, and
whose action on a natural transformation $\eta$ is the natural
transformation given by $H\,\eta = \id + \eta \times
F\,\eta$. Reflecting the functorial action of $\mu H$ back into syntax
gives exactly Haskell's built-in $\mathsf{map}$ function as the
type-and-data-uniform, structure-preserving, data-changing function
associated with $\mathsf{List}$. {\color{blue} Double-check this
  paragraph.}

In a parametric model, short cut fusion can be used to show that the
Church encoding and the primitive representation of an ADT are
isomorphic as sets. Note, however, that whereas it is impossible even
to {\em state} the rearrange-transform property for Church encodings
of ADTs, this property is simply the reflection back into syntax of
the instance of naturality for the type-independent function that
rearranges structures of type $\mathsf{T\,a}$ into ones of type
$\mathsf{T'\,a}$ and the functions $\mathsf{map_T\,f}$ and
$\mathsf{map_{T'}\,f}$ for a function $\mathsf{f :: a \to b}$ and the
type constructors $\mathsf{T}$ and $\mathsf{T'}$ associated with these
ADTs, respectively. {\color{blue} Double-check this paragraph.}

>>>>

GADTs are traditionally represented via Church encodings. In this
view, GADTs are completely determined by their syntax; for example,
the GADT $\mathsf{Seq}$ defined in Haskell by
\begin{equation}\label{eq:seq}
\begin{array}{l}
\mathsf{data\, Seq\,a\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Seq \,a \to Seq\,b \to
  Seq\,(a \times b)}\\ 
\end{array}
\end{equation}
\noindent
comprises sequences of any type $\mathsf{a}$ and sequences obtained by
pairing the data in two already existing sequences. By contrast, the
primitive representation sees a GADT as the {\em functorial
  completion}~\cite{jp19} of its syntax. Completing a GADT's syntax to
a {\em functor} is necessary in order to be able to take the least
fixed point~\cite{tfca}, so it is inherent in the view of data types
as their primitive representations. Viewed as its primitive
representation, $\mathsf{Seq}$ comprises not just those data elements
representable in syntax, but also all data elements of the form
$\mathsf{map\,f\,s}$ for all types $\mathsf{t_1}$ and $\mathsf{t_2}$,
all functions $\mathsf{f : t_1 \to t_2}$ definable in the language,
and all $\mathsf{s :: Seq\,t_1}$, as well as all data elements of the
form $\mathsf{map\,g\,s'}$ for each appropriately typed function
$\mathsf{g}$ and each element $\mathsf{s'}$ already added to the data
type, and so on. Functorial completion for $\mathsf{Seq}$ adds, in
particular, data elements of the form
$\mathsf{map\,g\,(Pair\,u_1\,u_2)}$ even though these may not
themselves be of the form $\mathsf{Pair\,v_1\,v_2}$ for any terms
$\mathsf{v_1}$ and $\mathsf{v_2}$. By contrast, functorial completion
adds no new data elements to the syntax of a GADT that is actually a
(polynomial) ADT or a (hereditarily polynomial) nested type, so its
Church encoding and primitive representation coincide. However, these
representations are rarely the same for GADTs.

@@

Requiring $F$ to be a functor will ensure that the interpretation $\mu
F$ of the ADT $\mathsf{T\,a} = \mathsf{\mu F}$ exists. But to ensure
that $\mu F$ is itself a functor, so that the type constructor
$\mathsf{T}$ associated with the ADT $\mathsf{T\,a}$ supports a
$\mathsf{map}$ function as well, we can require that $F$ also be the
fixpoint $\mu H$ of a functor $H$. In this case, however, $H$ must be
a functor on the category $\mathit{Set}^\mathit{Set}$ of functors and
natural transformations on $\mathit{Set}$. That is, $H$ must be a {\em
  higher-order} functor on $\mathit{Set}$. Reflecting this requirement
back into syntax gives that
$\mathsf{T\,a} = \mathsf{(\mu H)\,a}$
for the ``type constructor constructor''
$\mathsf{H}$ that supports suitable\footnote{The $\mathsf{map}$
  function for $\mathsf{H}$ should satisfy syntactic reflections of
  the functoriality properties in $\mathit{Set}^\mathit{Set}$ ---
  i.e., preservation of identity natural transformations and
  composition of natural transformations --- and the $\mathsf{map}$
  function for $\mathsf{H\,F}$ should satisfy syntactic reflections of
  the functoriality properties in $\mathit{Set}$ as in the previous
  footnote. Again, there is no mechanism in Haskell for enforcing
  this.} $\mathsf{map}$ functions.

@@

A concrete example is given by the ADT $\mathsf{List\,a}$. This type
is modelled as the fixpoint $\mu F_{\mathit{List\,a}}$ of the
first-order functor whose action on sets is given by
$F_{\mathit{List}\,a}\,X = 1 + a \times X$ and whose action on
functions is given by $F_{\mathit{List}\,a}\,f = 1 + \id_a \times
f$. The type constructor $\mathsf{List}$ is modelled by the functor
that is the fixpoint $\mu H$ of the higher-order functor $H$ whose
action on a functor $F$ is given by the functor $H\,F$ whose actions
on sets and functions between them are given by $H\,F\,X = 1 + X
\times F\,X$ and $H\,F\,f = \id_1 + f \times F\,f$, respectively, and
whose action on a natural transformation $\eta$ is the natural
transformation whose component at $X$ is given by $(H\,\eta)_X = \id_1
+ \id_X \times \eta_X$. Reflecting the functorial action of $\mu H$
back into syntax gives exactly Haskell's built-in $\mathsf{map}$
function as the type-and-data-uniform, structure-preserving,
data-changing function associated with $\mathsf{List}$.

In a parametric model, short cut fusion can be used to show that the
Church encoding and the primitive representation of an ADT are
isomorphic as sets. Note, however, that whereas it is impossible even
to {\em state} the rearrange-transform property for Church encodings
of ADTs, this property is simply the reflection back into syntax of
the instance of naturality for the type-independent function that
rearranges structures of type $\mathsf{T\,a}$ into ones of type
$\mathsf{T'\,a}$ and the content-transforming functions
$\mathsf{map_T\,f}$ and $\mathsf{map_{T'}\,f}$ for a function
$\mathsf{f :: a \to b}$, where $\mathsf{T}$ and $\mathsf{T'}$ are the
type constructors associated with these ADTs, respectively.
